CRUSE to Safe Cycling in Ireland \\\large{An Open Source Methodology to Support Active Travel}

### 2023-10-03 paper.qmd
Update abstract

---
title: "CRUSE to Safe Cycling in Ireland"
subtitle: An Open Source Methodology to Support Active Travel
bibliography: bibliography.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
number-sections: true

# Uncomment for draft versions:
format: 
  html: default
  docx: default
#   # pdf: default

# # Uncomment for formatted version
# format:
#   elsevier-pdf:
#     keep-tex: true
#     journal:
#       name: Journal Name
#       formatting: preprint
#       model: 3p
#       cite-style: super

@@

  To address data needs for both safety, business case development and project evaluation on the National Road Network (NRN), the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was developed.
  Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
  High resolution datasets on baseline and potential cycling levels are needed to meet Ireland's climate and safety targets.

  Due to an absence of reliable data, particularly rural cycling levels, TII commissioned the Institute for Transport Studies (ITS) at the University of Leeds and AECOM to develop a new tool for this purpose.
  ITS Leeds led the development of the PCT for England and Wales, which has "revolutionized the practice of strategic cycle planning nationally".
  The tool is an open-source approach using recognized open-source methodology to enable planners, engineers, and other stakeholders to make evidence-based decisions for the NRN.
  CRUSE is available at <https://cruse.bike/> and builds on the Propensity to Cycle Tool (PCT) for England and Wales.
  CRUSE goes beyond the PCT in several important ways, modeling denser networks at higher levels of geographic resolution and additional purposes, including simulated recreational, social, shopping and personal trips, in addition trips to work and primary, secondary and tertiary education locations.
  The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.

  CRUSE is structured in a similar way to the traditional four-stage transport model, but its use of Open Street Map (OSM) data, used by [Cyclestreets.net](https://www.cyclestreets.net/) for routing, enables network quality to be assessed without costly surveys to record new infrastructure.
  OSM tags generate "cycle friendliness" estimates of all links on the network, based on existing recorded infrastructure.
  Many networks are provided, highlighting routes for directness (Fastest) and "cycle friendliness" (Quietest).
  It uses origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.

  For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
  With growth in the E-bike market, the tool will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
  The findings and resulting tool are directly relevant to addressing the challenges and opportunities faced by National Road Authorities (NRAs) across Ireland.
  The provision of open access evidence, that can be used non-experts and professionals, is relevant to countries with sustainable transport policies worldwide.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
date: last-modified
---

>>>>

---
title: "CRUSE to Safe Cycling in Ireland"
subtitle: An Open Source Tool to Support Active Travel
bibliography: bibliography.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
number-sections: true

# # Uncomment for draft versions:
format: 
  html: default
  docx: default
# Uncomment for formatted version
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super
# #   # pdf: default

@@

  The Cycle Route Uptake and Scenario Estimation (CRUSE) Tool was commissioned by TII and developed by the Leeds Institute for Transport Studies (ITS) and AECOM To address data needs for safety, business case development and project evaluation on the National Road Network (NRN).
  Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
  High resolution datasets on baseline and potential cycling levels are needed to meet Ireland's climate and safety targets.

  The CRUSE project builds on the methods underlying the Propensity to Cycle Tool for England and Wales (PCT), which had a major impact on strategic cycle network planning in the UK.
  Like the PCT, CRUSE is based on open source software and reproducible data science. The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions for the NRN.
  CRUSE goes beyond the PCT in several important ways, modeling denser networks at higher levels of geographic resolution and additional purposes, including simulated recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
  The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.

  Another key feature of CRUSE is its use of Open Street Map (OSM) data, which is continuously updated, and used by [Cyclestreets.net](https://www.cyclestreets.net/) for route calculation and "cycle friendliness" estimates of all links on the network.
  CRUSE provides three network types: the 'Fastest' network highlights routes for directness; 'Quietest' prioritises cycle friendliness, and 'Balanced' represents a balance between speed and comfort.
  Trips are generated based on origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.

  For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
  With growth in the E-bike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
  Provision of open access data in a publicly available web tool will address the challenges faced by National Road Authorities (NRAs) across Ireland in meeting their sustainability commitments.
  The experience of developing a deploying CRUSE are relevant to countries with sustainable transport policies worldwide.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
date: last-modified
---

### 2023-10-03 paper.tex
Update abstract

To address data needs for both safety and project evaluation on the
National Road Network (NRN), the Cycle Route Uptake and Scenario
Estimation (CRUSE) Tool was developed. While cycling in Ireland
represents only 3\% of total modal share, with higher intensities in
urban areas, the levels of cycling collisions are disproportionately
high at 20\% of all serious injuries and 7\% of all fatalities. If
Ireland is to meet its climate and safety targets, data to establish
baseline cycling levels and future cycling levels is needed.

Due to an absence of reliable data, particularly rural cycling levels,
TII commissioned the Institute for Transport Studies (ITS) at the
University of Leeds and AECOM to develop a new tool for this purpose.
ITS Leeds led the development of the PCT for England and Wales, which
has ``revolutionized the practice of strategic cycle planning
nationally''. The tool is an open-source approach using recognized
open-source methodology to enable planners, engineers, and other
stakeholders to make evidence-based decisions for the NRN. CRUSE is
available at \url{https://cruse.bike/} and builds on the Propensity to
Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
several important ways, higher resolution data, more trip types,
including estimates for education, inter-urban, and recreational trips.
In addition to understanding cycling intensity, for asset planning and
management purposes, the tool provides essential road safety information
to enable reporting of disaggregate collision rates.

CRUSE is structured in a similar way to the traditional four-stage
transport model, but its use of Open Street Map (OSM) data, used by
\href{https://www.cyclestreets.net/}{Cyclestreets.net} for routing,
enables network quality to be assessed without costly surveys to record
new infrastructure. OSM tags generate ``cycle friendliness'' estimates
of all links on the network, based on existing recorded infrastructure.
A range of networks is provided, highlighting routes for directness
(Fastest) and ``cycle friendliness'' (Quietest). It uses origin and
destination data from the 2016 Census in combination with modeled demand
data to estimate cycling levels and potential at the area, route, and
network levels for each county in Ireland and offers estimates of the
baseline level of cycling and several future scenario-based levels of
cycling.

As countries, like Ireland, invest in cycling, the number of killed and
seriously injured cyclists must reduce too. The CRUSE Tool provides
estimates of cycling potential and routing for each county in Ireland,
and works in both urban and rural settings, to enable monitoring of
cycling safety. With growth in the E-bike market, the tool will help
inform inter-urban and rural networks to support the transfer of trips
to sustainable modes for longer journeys. The CRUSE Tool methodology and
findings are directly relevant to addressing the challenges and
opportunities faced by other NRAs. The datasets resulting from the
project are open access and can be used by both non-experts and
professionals.
\end{abstract}

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

Transport is responsible for 23\% of global emissions, 70\% of which is
from road transport, nearly half of which (around 10\% of global
emissions) is from passenger cars \citep{jaramillo2022}. The transport
system encourages, enables and in some cases enforces unsustainable
lifestyles, including over-consumption of goods due to excessive mobile
storage space and dependency on services that are only accessible by car
due to land use plans that have built up around roads
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The framework highlights the importance of demand
reduction (\emph{avoid}ing unnecessary trips), in addition to mode
\emph{shift} and \emph{improvement} of existing energy converters, in
that order.

Uptake of cycling, the main topic of this paper, should be seen in this
broader context of transport decarbonisation \citep{brand2020} and
sustainable mobility \citep{burns2013}. Although cycling uptake appears
on the surface to only relate to the `shift' part of the ASI framework,
closer consideration of the knock-on impacts of cycling uptake shows
that it can also help avoid unnecessary trips \citep{nello-deakin2020}.
The rapid uptake of highly efficient e-bikes can also be seen as an
improvement on most electric vehicles, which are too heavy and expensive
to be a sustainable alternative and could in fact delay the transition
away from car dependency and inadvertently enable ``high travel
lock-in'' \citep{anable2019}. At the European level, the European Union
has a target of reducing greenhouse gas emissions by 55\% by 2030,
compared to 1990 levels, and to achieve `net-zero' by 2050.{[}\^{}1{]}

Regarding road traffic casualties, another deadly consequence of
inefficient transport systems, the Road Infrastructure Safety Management
(RISM) directive (2008/96/EC) requires member states to implement a road
safety management system (RSMS) for all public roads. Specifically,
``Member States shall ensure that the ranking of high accident
concentration sections and the network safety ranking are carried
out''.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
  ri=celex\%3A32008L0096} Given that `safety' in this context is
usefully quantified as the number of people killed and seriously injured
(KSI) per distance travelled, the directive requires estimation of
distance travelled by mode, down to the road link level. For active
modes, about which there is a paucity of data compared with motorized
modes, this is a major challenge. Better data to inform road safety
policies and interventions is a motivation for the CRUSE tool outlined
in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by
2030.{[}\^{}3{]} Doing so while simultaneously enabling rapid uptake of
active modes will require key active travel routes to be identified and
improved.

A proactive approach to the development of cycling policy and
infrastructure is taking place in Ireland. The main framework
underpinning these efforts is the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}}
and The National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
. Strategic alignment is also outlined by the Phoenix Park Transport and
Mobility Options Study, which highlights public support for active
travel interventions: ``83\% of respondents supported enhanced walking
and cycling facilities'' .

At the regional level, the recently published Greater Dublin Area
Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
reinforces these findings: ``nearly a quarter of adults cycle at least
once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
area taking up to 60,000 cars off the road today. Extrapolating this on
a per population basis across Ireland, with around 40\% of the
population living in Dublin, this suggests that around 150,000 cars
could be removed nationwide just by achieving Dublin levels of cycling
in all counties (notwithstanding existing cycling trips and differences
in trip distances).

@@

It was in this context that Transport Infrastructure Ireland (TII)
commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE)
project. Based on the Propensity to Cycle Tool (PCT) for England and
Wales, the CRUSE tool was developed to provide evidence on current
cycling levels and future cycling potential nationwide across Ireland. A
key consideration for TII is to ensure both urban and rural Ireland were
targeted for cycling infrastructure, as shown in
Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
on Irish Roads and how the road network was being used by commuters,
cyclists, heavy goods vehicles and other road users across the Irish
transport network. The aim was for the tool to provide strong, national,
systematic but locally-specific evidence to monitor cycling friendliness
and safety in Ireland, to ensure strategic alignment with national,
regional and local policies. The open source and publicly available
nature of the tool ensures more inclusive and evidence-based
conversations around cycle network planning between all stakeholders in
the transport planning process. Other use cases in the Irish context
includes monitoring the need for existing cycling infrastructure
upgrades, undertaking appraisal, as well as safety and exposure
information in line with the European Cycling Federation, Dutch Cycling
Embassy recommendations and the European RISM Directive, for reporting
collision rates of vulnerable road users by 2024. These factors mean
that the tool can be seen as an open access `leverage point' in the
planning system \citep{lovelace2020}.

@@

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland. Source: CRUSE tool,
available at \url{https://cruse.bike/}.}

@@

The rest of paper is structured as follows. In
Section~\ref{sec-methods}, we outline the methods used to generate the
evidence presented in the CRUSE tool for Ireland. In
Section~\ref{sec-results}, we present the results of the CRUSE tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss the implications of the results
for policy and practice. In Section~\ref{sec-conclusions}, we conclude
the paper.

@@

The methods used to generate the evidence presented in the CRUSE tool
for Ireland build on the Propensity to Cycle Tool (PCT), which was
originally funded by the UK's Department for Transport and developed by
a multi-university team. An important feature of the PCT is that it is
open source and publicly available (at
\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
stakeholders in the transport planning process \citep{lovelace2017}. The
PCT approach has had major policy impacts, as outlined in Research
Excellence Framework (REF) impact case studies, which demonstrate that
the tool ``revolutionised strategic cycle planning in England and
Wales''\footnote{See REF Impact Case Study ``Cycle network policy,
  planning and investment transformed by the Propensity to Cycle Tool''
  at
  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
  submitted by the University of Leeds.} by overcoming the barriers to
cycling investment imposed by lack of evidence on cycling
potential\footnote{See the REF Impact Case Study ``Creating Step Changes
  in Cycling Policy and Infrastructure Planning across the UK'' at
  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
  submitted by the University of Westminster.} .

@@

However, travel to school and work only constitute a small proportion of
all trips. To overcome this issue, we developed a spatial interaction
modelling methodology to estimate the number of trips between each OD
pair for additional trip purposes. The classification of trip purposes
used in the CRUSE tool was guided mainly by the trip purpose
classification found within the National Household Travel Survey (NHTS),
but with the addition of categories based on the comprehensive POWSCAR
data, and the need to include recreational trips and multi-stage trips
(not yet implemented). An overview of the trip purposes used in CRUSE is
presented in Figure~\ref{fig-trip-purposes}.

@@

\caption{\label{fig-trip-purposes}Trip purposes used in the CRUSE tool.}

@@

\caption{\label{fig-user-stories}User stories for the CRUSE tool}

@@

National planning

Regional

Local

@@

>>>>

To address data needs for both safety, business case development and
project evaluation on the National Road Network (NRN), the Cycle Route
Uptake and Scenario Estimation (CRUSE) Tool was developed. Cycling in
Ireland represents only 3\% of total modal share as of the latest Census
(2016), but accounts for 20\% of serious injuries and 7\% of all
fatalities. High resolution datasets on baseline and potential cycling
levels are needed to meet Ireland's climate and safety targets.

Due to an absence of reliable data, particularly rural cycling levels,
TII commissioned the Institute for Transport Studies (ITS) at the
University of Leeds and AECOM to develop a new tool for this purpose.
ITS Leeds led the development of the PCT for England and Wales, which
has ``revolutionized the practice of strategic cycle planning
nationally''. The tool is an open-source approach using recognized
open-source methodology to enable planners, engineers, and other
stakeholders to make evidence-based decisions for the NRN. CRUSE is
available at \url{https://cruse.bike/} and builds on the Propensity to
Cycle Tool (PCT) for England and Wales. CRUSE goes beyond the PCT in
several important ways, modeling denser networks at higher levels of
geographic resolution and additional purposes, including simulated
recreational, social, shopping and personal trips, in addition trips to
work and primary, secondary and tertiary education locations. The tool
provides key `denominator' information to enable reporting of collision
rates per billion km cycled.

CRUSE is structured in a similar way to the traditional four-stage
transport model, but its use of Open Street Map (OSM) data, used by
\href{https://www.cyclestreets.net/}{Cyclestreets.net} for routing,
enables network quality to be assessed without costly surveys to record
new infrastructure. OSM tags generate ``cycle friendliness'' estimates
of all links on the network, based on existing recorded infrastructure.
Many networks are provided, highlighting routes for directness (Fastest)
and ``cycle friendliness'' (Quietest). It uses origin and destination
data from the 2016 Census in combination with modeled demand data to
estimate current and potential future cycling levels at area, route, and
network levels for each county in Ireland.

For investment in cycling to deliver on the huge potential benefits,
cycling must become safer. With growth in the E-bike market, the tool
will help inform inter-urban and rural networks to support the transfer
of trips to sustainable modes for longer journeys. The findings and
resulting tool are directly relevant to addressing the challenges and
opportunities faced by National Road Authorities (NRAs) across Ireland.
The provision of open access evidence, that can be used non-experts and
professionals, is relevant to countries with sustainable transport
policies worldwide.
\end{abstract}

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

Transport is responsible for 23\% of global greenhouse gas emissions,
70\% of which are from road transport, nearly half of which (around 10\%
of global emissions) is from passenger cars \citep{jaramillo2022}. The
transport system encourages, enables and in some cases enforces
unsustainable lifestyles. Services that are only accessible by car
lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The framework highlights the importance of demand
reduction (\emph{avoid}ing unnecessary trips), in addition to mode
\emph{shift} to sustainable modes and \emph{improvement} of existing
energy converters, in that order.

Uptake of cycling, the main topic of this paper, should be seen in this
broader context of transport decarbonisation \citep{brand2020} and
sustainable mobility \citep{burns2013}. Although cycling uptake appears
on the surface to only relate to the `shift' part of the ASI framework,
closer consideration of the knock-on impacts of cycling uptake shows
that it can also help avoid unnecessary trips \citep{nello-deakin2020}.
Furthermore, highly efficient e-bikes --- which are seeing rapid uptake
in Ireland and other countries --- outperform electric cars, which are
too heavy and expensive. Over-reliance on electric cars could slow the
transition away from car dependency and inadvertently enable ``high
travel lock-in'' \citep{anable2019}. At the European level, the European
Union has a target of reducing greenhouse gas emissions by 55\% by 2030,
compared to 1990 levels, and to achieve `net-zero' by 2050.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
  ri=celex\%3A32008L0096}

Regarding road traffic casualties, another deadly consequence of
inefficient transport systems, the Road Infrastructure Safety Management
(RISM) directive (2008/96/EC) requires member states to implement a road
safety management system (RSMS) for all public roads. Specifically,
``Member States shall ensure that the ranking of high accident
concentration sections and the network safety ranking are carried
out''.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
  ri=celex\%3A32008L0096} Given that `safety' in this context is
usefully quantified as the number of people killed and seriously injured
(KSI) per distance travelled, the directive requires estimation of
distance travelled by mode, down to the road link level. For active
modes, about which there is a paucity of data compared with motorized
modes, this is a major challenge. Better data to inform road safety
policies and interventions is a motivation for the CRUSE Tool outlined
in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by
2030.\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
Doing so while simultaneously enabling rapid uptake of active modes will
require key active travel routes to be identified and improved.

A proactive approach to the development of cycling policy and
infrastructure is taking place in Ireland. The main frameworks
underpinning these efforts are the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}},
the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
and the National Roads 2040 strategy\footnote{
  https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
These policy frameworks are backed-up by public support for active
travel interventions: while only 3\% of trips in Ireland are made by
cycling according to the latest Census data from 2016,\footnote{
  https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
``83\% of respondents supported enhanced walking and cycling
facilities''.

At the regional level, the recently published Greater Dublin Area
Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
reinforces these findings: ``nearly a quarter of adults cycle at least
once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
area taking up to 60,000 cars off the road today. Extrapolating this on
a per population basis across Ireland, with around 40\% of the
population living in Dublin, this suggests that around 150,000 cars
could be removed nationwide just by achieving Dublin levels of cycling
in all counties (notwithstanding existing cycling trips and differences
in trip distances). 7 in 10 trips in Ireland are by car\footnote{
  https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}.
Cycling has the potential to replace a large proportion of these trips:
``A high priority must also be given to cyclists, because trips by this
mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as e-bikes extend the range of this mode''\footnote{
  https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.

@@

It was in this context that Transport Infrastructure Ireland (TII)
commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE)
project. Based on the Propensity to Cycle Tool (PCT) for England and
Wales, the CRUSE Tool was developed to provide evidence on current
cycling levels and future cycling potential nationwide across Ireland.

A key consideration for TII is to ensure both urban and rural Ireland
were targeted for cycling infrastructure, as shown in
Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
on Irish Roads and how the road network was being used by commuters,
cyclists, heavy goods vehicles and other road users across the Irish
transport network. The aim was for the tool to provide strong, national,
systematic but locally-specific evidence to monitor cycling friendliness
and safety in Ireland, to ensure strategic alignment with national,
regional and local policies. The open source and publicly available
nature of the tool ensures more inclusive and evidence-based
conversations around cycle network planning between all stakeholders in
the transport planning process. Other use cases in the Irish context
includes monitoring the need for existing cycling infrastructure
upgrades, undertaking appraisal for business case development, as well
as safety and exposure information. The latter is in line with the
European Cycling Federation, Dutch Cycling Embassy recommendations and
the European RISM Directive, for reporting collision rates of vulnerable
road users by 2024. These factors mean that the tool can be seen as an
open access `leverage point' in the planning system
\citep{lovelace2020}.

@@

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE Tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland. Source: CRUSE Tool,
available at \url{https://cruse.bike/}.}

@@

The rest of paper is structured as follows. In
Section~\ref{sec-methods}, we outline the methods used to generate the
evidence presented in the CRUSE Tool for Ireland. In
Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss the implications of the results
for policy and practice. In Section~\ref{sec-conclusions}, we conclude
the paper.

@@

The methods used to generate the evidence presented in the CRUSE Tool
for Ireland build on the Propensity to Cycle Tool (PCT), which was
originally funded by the UK's Department for Transport and developed by
a multi-university team. An important feature of the PCT is that it is
open source and publicly available (at
\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
stakeholders in the transport planning process \citep{lovelace2017}. The
PCT approach has had major policy impacts, as outlined in Research
Excellence Framework (REF) impact case studies, which demonstrate that
the tool ``revolutionised strategic cycle planning in England and
Wales''\footnote{See REF Impact Case Study ``Cycle network policy,
  planning and investment transformed by the Propensity to Cycle Tool''
  at
  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
  submitted by the University of Leeds.} by overcoming the barriers to
cycling investment imposed by lack of evidence on cycling
potential\footnote{See the REF Impact Case Study ``Creating Step Changes
  in Cycling Policy and Infrastructure Planning across the UK'' at
  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
  submitted by the University of Westminster.} .

@@

However, travel to school and work only constitute around only 30\% of
all trips in Ireland. To overcome this issue, we developed a spatial
interaction modelling methodology to estimate the number of trips
between each OD pair for additional trip purposes. The classification of
trip purposes used in the CRUSE Tool was guided mainly by the trip
purpose classification found within the National Household Travel Survey
(NHTS), but with the addition of categories based on the comprehensive
POWSCAR data, and the need to include recreational trips and multi-stage
trips (not yet implemented). An overview of the trip purposes used in
CRUSE is presented in Figure~\ref{fig-trip-purposes}.

@@

\caption{\label{fig-trip-purposes}Trip purposes used in the CRUSE Tool.}

@@

\caption{\label{fig-user-stories}Hypothetical user stories for the CRUSE
Tool illustrating that it can be used in different ways by different
user groups.}

@@

The main result of the CRUSE project is an open access web application
and datasets on current and potential future cycling levels in Ireland,
with evidence provided at county and route segment levels. Like the PCT
on which CRUSE builds, a key feature of the results is that they are
systematic and available for all counties, and most routes on which
cycling is permitted, in Ireland. The full set of results are therefore
too extensive to present in their entirety in this paper. Through the
interactive web application, users can explore the data to generate the
results that are most relevant to their needs, whether that is finding
the cycling potential on a particular road or finding `weak links' or
barriers in the cycle network associated with a particular school, work
place or other destination.

Instead of trying to show such use cases, of which there are many
hundreds, we present a selection of results to illustrate the main
features of the CRUSE Tool.

To illustrate the results in urban areas, we took the top 4 cities in
Ireland by population: Dublin City, Cork, Limerick and Galway (with city
populations of 0.6, 0.2, 0.1 and 0.1 million respectively). The results
for central areas demarcated with circular areas with a radius of 3km
for these cities are presented in Figure~\ref{fig-city-results}.

\includegraphics{images/rnet_types.png}

\caption{\label{fig-city-results}Fastest, balanced and quietest cycle
networks within 3km circular areas containing the four largest cities in
Ireland. The networks are coloured by cycle friendliness and with widths
varying in proportion to the potential level of cycling under the `Go
Dutch' scenario.}

Aside from the widely varying sizes and route networks of each city,
with Dublin having by far the densest network, it is clear from the
visualisation of cycling friendliness that there is a lot of room for
improvement in all cities. Dublin has had received most investment in
cycling infrastructure and this is apparent from the high proportion of
the route network that is green, even in the `Fastest' route network
representing routes that prioritise speed and directness over comfort
and safety. Dublin also has the highest mode share of cycling of the
four cities, adding to the evidence base showing that investment in
cycling infrastructure is effective in increasing cycling uptake.

Although Dublin City has made progress, there are still many parts of
the Fastest route network, and even some parts of the Balanced and
Quietest route networks, that are not cycle friendly and which have high
cycling potential. According to a recent report, 71\% of residents in
the Dublin Metropolitan Area support ``more cycle tracks along roads,
physically separated from traffic and pedestrians''.\footnote{https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22\_DublinMetropolitanArea\_v35\_DIGITAL\_v2.pdf}
The results for the Dublin area could help prioritise investment in
cycling infrastructure in the city.

@@

The CRUSE Tool provides evidence on current and potential future cycling
levels across Ireland down to the street level, with potential assigned
to `fastest', `balanced' and `quietest' route networks. This provision
of multiple scenarios of behaviour changes \emph{and} multiple scenarios
of investment, for example in cycling infrastructure next to major roads
vs quiet residential streets, is a key feature of the CRUSE Tool.

As highlighted in a paper on cycling infrastructure preferences based on
a case study of Dublin, both directness and quietness are important
\citep{caulfield2012}:

\begin{quote}
Direct routes with short journey times were found to be the most
important positive variable for existing cyclists and non-cyclists in
determining route choice. This is followed by infrastructure type, the
number of junctions along the route, traffic speed and cyclist volumes.
In terms if infrastructure, regardless of the level of cycling
confidence, routes which have `no facilities' or `bus/cycle lanes' are
the least favoured cycle route types.
\end{quote}

The CRUSE Tool can help both in terms of describing the current
situation, and also prioritise investment in those direct routes with
high cycling potential that lack adequate cycling infrastructure,
highlighted by \citet{caulfield2012}. It will also support reporting for
the RISM Directive on road casualties as part of the road safety
management system.

While the CRUSE Tool is not the first national and publicly available
tool for cycling planning, it has some unique features that make it
relevant for other countries, and other national road authorities tasked
with making their transport systems safe.

TODO: outline these features.

The CRUSE Tool is designed to be used by a wide range of stakeholders,
including transport planners, engineers, and other professionals, as
well as non-professionals.
